Sequential rank agreement methods for comparison of lists

### 2015-07-14 latex/paper.Rnw
Discussion

The manuscript is organized as follows: In the next section we define
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete/censored lists. In section 3 we present and discuss
approaches to evaluate the sequential rank agreement obtained from an
experiment and to identify the maximal depth until which the ranked
lists agree to certain extent. Finally we present three different
applications of the proposed sequential rank agreement approach to
real data before we discuss the findings along with possible
extensions.

@@

<<example1, echo=FALSE, eval=FALSE>>=
library(SuperRanker)
library(glmnet)
library(minerva)
library(changepoint)
library(randomForestSRC)
# Read data
library(multtest)
data(golub)

@@

<<fig1a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE,eval=FALSE>>=
ysize <- 1000
nreference <- 20
## Plot the line
plot(res[1:ysize], col="black", lwd=3, type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<fig1b, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE, eval=FALSE>>=
## Same as above but zoomed in
ysize <- 50
plot(res[1:ysize], lwd=3, col="black", type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<label=tab1,echo=FALSE,results=tex, eval=FALSE>>=
library(xtable)
xxx <- as.data.frame(cbind(1:10, inputmatrix[1:10,]))
colnames(xxx) <- c("Ranking", colnames(xxx)[2:5])
xxx2 <- xtable(xxx, caption = "Top 10 list of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method")
digits(xxx2) <- 0
print(xxx2, table.placement = "tb", caption.placement = "top", include.rownames=FALSE)
#print(xtable(xxx,  label = "tab:one",

@@

%% \begin{figure}[tb]
%% \begin{center}
%% \includegraphics[width=.45\textwidth]{paper-fig-ao1}
%% \includegraphics[width=.45\textwidth]{paper-fig-ao2}
%% \end{center}
 %% \caption{Average overlap (left figure) and sequential rank agreement
   %% (right figure) for comparing two different analysis methods
   %% (marginal $t$ test and logistic regression)
   %% applied to the Golub data. The black lines are based on the full
   %% set of predictors while the red lines have the two highest
   %% associated predictors removed
   %% from the data before computing the average overlap and sequential
   %% rank agreement.}
 %% \label{fig:case1}
%% \end{figure}

The black lines Figure~\ref{fig:case1} show the average overlap and
sequential rank agreement for these data. It is clear from both plots
that there is perfect agreement towards the top of the ranked lists:
the average overlap is 1 and the sequential rank agreement is zero.
However, if we remove those two items from the data (i.e.,
gene/predictor 2124 and 896) and redo the analyses then we get
substantially different curves for the average overlap but not for
sequential rank agreement (red lines in Figure~\ref{fig:case1}). For
the sequential rank agreement we get roughly the same estimate of
agreement from rank 3 as we did from the full dataset. This indicates
that sequential rank is more robust against small pertubations of the
data. For the average overlap the curves completely change when these
two items are removed and which would lead to a different conclusion
about the agreement among the items from rank 3 downwards.

@@

\section{Discussion}

Mention/discuss different measures.

Censor lists at significane level?

Bootstrap across a single method and compare results. Discuss collinearity

Changepoint analysis

\bibliographystyle{rss}
\bibliography{paperref}

>>>>

The manuscript is organized as follows: In the next section we define
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete/censored lists. In section 3 we present and discuss
approaches to evaluate the sequential rank agreement obtained from an
experiment and to identify the maximal depth until which the ranked
lists agree to certain extent. Finally we present three different
applications of the proposed sequential rank agreement approach to
real data before we discuss the findings along with possible
extensions. The approaches presented in this manuscript are available
in the R package \texttt{SuperRanker}.

@@

<<example1, echo=FALSE, eval=TRUE>>=
library(SuperRanker)
library(glmnet)
library(minerva)
library(changepoint)
library(randomForestSRC)
# Read data
library(multtest)
data(golub)

@@

<<fig1a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE,eval=TRUE>>=
ysize <- 1000
nreference <- 20
## Plot the line
plot(res[1:ysize], col="black", lwd=3, type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<fig1b, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE, eval=TRUE>>=
## Same as above but zoomed in
ysize <- 50
plot(res[1:ysize], lwd=3, col="black", type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<label=tab1,echo=FALSE,results=tex, eval=TRUE>>=
library(xtable)
xxx <- as.data.frame(cbind(1:10, inputmatrix[1:10,]))
colnames(xxx) <- c("Ranking", colnames(xxx)[2:5])
xxx2 <- xtable(xxx, caption = "Top 10 list of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method", label="tab1")
digits(xxx2) <- 0
print(xxx2, table.placement = "tb", caption.placement = "top", include.rownames=FALSE)
#print(xtable(xxx,  label = "tab:one",

@@

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.45\textwidth]{paper-fig-ao1}
\includegraphics[width=.45\textwidth]{paper-fig-ao2}
\end{center}
 \caption{Average overlap (left figure) and sequential rank agreement
 (right figure) for comparing two different analysis methods
 (marginal $t$ test and logistic regression)
    applied to the Golub data. The black lines are based on the full
    set of predictors while the red lines have the two highest
    associated predictors removed
    from the data before computing the average overlap and sequential
    rank agreement.}
  \label{fig:case1}
 \end{figure}

The black lines in Figure~\ref{fig:case1} show the average overlap and
sequential rank agreement for these data. It is clear from both plots
that there is perfect agreement towards the top of the ranked lists:
the average overlap is 1 and the sequential rank agreement is zero.
However, if we remove those two items from the data (i.e.,
gene/predictor 2124 and 896) and redo the analyses then we get
substantially different curves for the average overlap but not for
sequential rank agreement (red lines in Figure~\ref{fig:case1}). For
the sequential rank agreement we get roughly the same estimate of
agreement from rank 3 as we did from the full dataset. This indicates
that sequential rank is more robust against small pertubations of the
data. For the average overlap the curves completely change when these
two items are removed and which would lead to a different conclusion
about the agreement among the items from rank 3 downwards.

@@

\section{Discussion}
In this article we address the problem of comparing ranked lists of
the same items. Our proposed method can handle both the situation
where the underlying data to generate the ranked lists are available
and the situation where the only available data is the actual ranked
lists. In addition, censored ranked lists where only the top $k$
ranked items are known can be accommodated as well. The proposed
agreement measure can be interpreted as the average distance between
an item's rank and the average rank assigned to that item and thus
determine how well the lists agree on the ranks of a specific
item and it circumvents the problem with other existing similarity measures.

The sequential rank agreement is extremely versatile. We have shown
that it can be used not only to compare ranked lists of items produced
from different samples/populations but that it also can be used to
study the ranks obtained from different analysis methods on the same
data as well as to evalaute the stability of the ranks from a single
method by bootstrapping the data repeatedly and comparing the ranks
obtained from the bootstrapped data.

Our applications show that XXX
Through simulations we showed that XXX --- afventer Andreas.

The sequential rank agreement can be used to determine the depth at
which the rank agreement becomes too large to be desirable based on
prior requirements or acceptable differences, or it can be used to
visually determine when the change in agreement becomes too large.  In
that regard the investigator can have prior limits on the level of
agreement that is acceptable.

While the sequential rank agreement is primarily an exploratory tool
we have suggested two null hypotheses that can be used to evaluate the
sequential rank agreement obtained. Note that none of the two null
hypotheses are concerned with the actual ``true ranking'' but are
purely concerned with consistency/stability of the rankings among the
lists. As such we cannot determine if the rankings are good but only
whether they agree. The sequential rank agreemenet curve can be
compared visually to the curves obtained under either of the null
distributions and simple point-wide $p$-values can be obtained for
each depth by counting the number of sequential rank agreements under
the null hypothesis that is less than or equal to the observed rank
agreement. Two simple extensions can be pursued to make more formal
uniform tests: one would be to use a Kolmogorov-Smirnov-like test and
use the largest difference between the variance-weighted sequential
rank agreement curve and the mean null rank agreement curve as a test
statistic. Alternatively, a change point analysis could be made on the
sequential rank agreement in order to determine the depths at which
there are ``jumps'' in the rank agreement. These jumps would
correspond to depths for which the agreement among the lists was
substantially worse and could serve as indicators for when the lists
no longer agree sufficiently satisfactory.

Finally, we have --- whenever possible --- used as much information
from the lists as possible, i.e., we have used all available ranks. If
we have information on the rankings as well as the corresponding
evidence for significance that we could choose to censor the rankings
of all the items that were deemed not to be significant. That would
ensure that there would be put less emphasis on the agreement of the
non-significant items and it would be easier to identify a change in
agreement among the items that were deemed to relevant.

We note that the sequential rank agreement is still marred by problems
that generally apply to ranking of items and/or
indiviuals. Collinearity in particular can be a huge problem when
bootstrapping data or when comparing different analysis methods. For
example, marginal analyses where each item is analyzed seperately will
assign similar ranks to two highly correlated predictors while methods
that provide a sparse solution such as the lasso will just rank one of
the two predictors high while the other might have a very low
rank. When the rankings from these two analysis methods are compared
then the agreement will differ simply because of the way correlated
predictors are handled. This is not a shortcoming of the sequential
rank agreement but is a problem general to all ranked lists.

In conclusion we have introduced a method for evaluation of ranked
(censored) lists that can be easily interpreted and that can be
applied to a large number of situations.  The method can be adapted
even further by using it to identify and cluster statistical analysis
methods that provide more or less the same agreement, or by using the
rank agreement as a hyper-parameter in, say, elastic net regularized
regression where the rank agreement is used to determine the mixing
proportion between the $L_1$ and the $L_2$ penalty. Finally, the
proposed method may be adapted (with some additional assumptions) to
the situation where the lists are subsets of each other. We will be
investigating these extensions further in the future.

%\bibliographystyle{rss}
\bibliographystyle{rss}
\bibliography{paperref}
